\vspace{-\baselineskip} Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-3.2\baselineskip}

### 2017-03-25 paper/ms.tex
words about rotation period uncertainties

GP are stochastic models consisting of a mean function
$\mu_\bvec{\theta}(\bvec{x})$ and a covariance, autocorrelation, or ``kernel''
function $k_\bvec{\alpha}(\bvec{x}_i,\,\bvec{x}_j)$ parameterized by the
parameters $\bvec{\theta}$ and $\bvec{\alpha}$ respectively.
Under this model, the log-likelihood of observing a dataset
\begin{eqnarray}
\bvec{y} = \left(\begin{array}{ccccc}
    y_1\quad && \cdots\quad && y_N
\end{array}\right)^\T
\end{eqnarray}
at coordinates
\begin{eqnarray}
X = \left(\begin{array}{ccccc}
    \bvec{x}_1\quad && \cdots\quad && \bvec{x}_N
\end{array}\right)^\T
\end{eqnarray}
is
\begin{eqnarray}\eqlabel{gp-likelihood}
\ln{p(\bvec{y}\,|\,{X,\,\bvec{\theta}},\,\bvec{\alpha})} =
    -\frac{1}{2} {\bvec{r}_\bvec{\theta}}^\T\,{K_\bvec{\alpha}}^{-1}\,
        \bvec{r}_\bvec{\theta}
    -\frac{1}{2}\ln\det K_\bvec{\alpha}
    - \frac{N}{2} \ln{(2\pi)}
\end{eqnarray}
where
\begin{eqnarray}
    \bvec{r}_\bvec{\theta} = \left(\begin{array}{ccccc}
    y_1 - \mu_\bvec{\theta}(\bvec{x}_1)\quad && \cdots\quad &&
    y_N - \mu_\bvec{\theta}(\bvec{x}_N)
\end{array}\right)^\T
\end{eqnarray}
is the vector of residuals and the elements of the covariance matrix $K$ are
given by $[K_\bvec{\alpha}]_{nm} = k_\bvec{\alpha}(\bvec{x}_n,\,\bvec{x}_m)$.
The maximum likelihood values for the parameters $\bvec{\theta}$ and
$\bvec{\alpha}$ for a given dataset $(\bvec{y},\,X)$ can be found by
maximizing \eq{gp-likelihood} with respect to $\bvec{\theta}$ and
$\bvec{\alpha}$ using a non-linear optimization routine \citep{Nocedal:2006}.
Similarly, probabilistic constraints on $\bvec{\theta}$ and $\bvec{\alpha}$
can be obtained by multiplying the likelihood by a prior
$p(\bvec{\theta},\,\bvec{\alpha})$ and using a Markov Chain Monte Carlo (MCMC)
algorithm to sample the joint posterior probability density.

@@

By rewriting the exponentials in \eq{celerite-kernel-complex} as sums of sine
and cosine functions, we can see the autocorrelation structure is defined by a
mixture of quasiperiodic oscillators
\begin{eqnarray}\eqlabel{celerite-kernel}
k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} +
    \sum_{j=1}^J &&\left[
    a_j\,\exp\left(-c_j\,\tau_{nm}\right)\,\cos\left(d_j\,\tau_{nm}\right)
        \right.\nonumber\\
    &&+ \left.
    b_j\,\exp\left(-c_j\,\tau_{nm}\right)\,\sin\left(d_j\,\tau_{nm}\right)
\right] \quad.
\end{eqnarray}
For clarity, we refer to the argument within the sum as a ``\celeriteterm\
term'' for the remainder of this paper.
The kernel function in \eq{celerite-kernel} is qualitatively similar to the
The Fourier transform\footnote{Here and throughout we have defined the Fourier
transform of the function $f(t)$ as $F(\omega)={(2\,\pi)}^{-1/2}\,
\int_{-\infty}^\infty f(t)\,e^{i\,\omega\,t}\dd t$.} of this covariance
function is the power spectral density (PSD) of the process and it is given by
\begin{eqnarray}\eqlabel{celerite-psd}
S(\omega) = \sum_{j=1}^J \sqrt{\frac{2}{\pi}}
\frac{(a_j\,c_j+b_j\,d_j)\,({c_j}^2+{d_j}^2)+(a_j\,c_j-b_j\,d_j)\,\omega^2}
{\omega^4+2\,({c_j}^2-{d_j}^2)\,\omega^2+{({c_j}^2+{d_j}^2)}^2}\quad.
\end{eqnarray}
The physical interpretation of this model isn't immediately obvious and we
return to a more general discussion of the physical intuition in a moment
but we start by investigating some special cases.

@@

Another source of variability that can be measured from time series
measurements of stars is rotation.
The inhomogeneous surface of the star (spots, plage, \etc) imprints itself as
quasiperiodic variations in photometric or spectroscopic observations
\citep{Dumusque:2014}.
It has been demonstrated that for light curves with nearly uniform sampling,
the empirical autocorrelation function provides a reliable estimate of the
rotation period of a star \citep{Mcquillan:2013, Mcquillan:2014, Aigrain:2015}
and that a GP model with a
quasiperiodic covariance function can be used to make probabilistic
measurements even with sparsely sampled data (R.~Angus, \etal\ in prep.).
The covariance function used for this type of analysis has the form
\begin{eqnarray}\eqlabel{sine2}
k(\tau) = A\,\exp\left(-\frac{\tau^2}{2\,\ell^2} -
    \Gamma\,\sin^2\left(\frac{\pi\,\tau}{P} \right) \right)
\end{eqnarray}
where $P$ is the rotation period of the star.
GP modeling with the same kernel function has been proposed as a method of
measuring the mean periodicity in quasiperiodic photometric time series in
general \citep{Wang:2012}.
The key difference between \eq{sine2} and other quasiperiodic kernels is that
it is positive everywhere.
We can construct a simple \celeriteterm\ covariance function with similar properties
as follows
\begin{eqnarray}\eqlabel{rot-kernel}
k(\tau) = \frac{a}{2+b}\,e^{-c\,\tau}\,\left[
    \cos\left(\frac{2\,\pi\,\tau}{P}\right) + (1 + b)
\right]
\end{eqnarray}
for $a>0$, $b>0$, and $c>0$.
The covariance function in \eq{rot-kernel} cannot exactly reproduce \eq{sine2}
but, since \eq{sine2} is only an effective model, \eq{rot-kernel} can be used
as a drop-in replacement for a substantial gain in computational efficiency.

To demonstrate this method, we fit a \celeriteterm\ model with a kernel given by
\eq{rot-kernel} to a \kepler\ light curve for the star KIC~1430163.
This star has a published rotation period of $3.88 \pm 0.58$, measured using
traditional periodogram and autocorrelation function approaches applied to
\kepler\ data from Quarters 0--16 \citep{Mathur:2014}.
We used \emcee\ to sample the joint posterior probability density for the four
parameters in \eq{rot-kernel}, conditioned on two quarters of \kepler\ data,
and find a constraint on the rotation period of $P = \rotationperiod$.
This is in good agreement with the literature value.
\Figure{rotation} shows a subset of the data used in this example and the
posterior inferences of the PSD and autocorrelation function of the process.
\Figure{rotation-period} shows the marginalized posterior distribution for the
rotation period.

@@

\newpage
\section{Comparisons to other methods}

@@

For a GP kernel to be valid, it must produce a positive definite covariance
matrix for all input coordinates.
For stationary kernels, this is equivalent~--~by Bochner's theorem \citep[see
Section 4.2.1 in][]{Rasmussen:2006}~--~to requiring that the kernel be the
Fourier transform of a positive finite measure.
This means that the power spectrum of a positive definite kernel must be
positive for all frequencies.
This result is also intuitive because, since the power spectrum of a process
is defined as expected the squared amplitude of the Fourier transform of the
time series, it must be non-negative.

>>>>

GP are stochastic models consisting of a mean function
$\mu_\bvec{\theta}(\bvec{x})$ and a covariance, autocorrelation, or ``kernel''
function $k_\bvec{\alpha}(\bvec{x}_i,\,\bvec{x}_j)$ parameterized by
$\bvec{\theta}$ and $\bvec{\alpha}$ respectively.
Under this model, the log-likelihood of observing a dataset
\begin{eqnarray}
\bvec{y} = \left(\begin{array}{ccccc}
    y_1\quad && \cdots\quad && y_N
\end{array}\right)^\T
\end{eqnarray}
at coordinates
\begin{eqnarray}
X = \left(\begin{array}{ccccc}
    \bvec{x}_1\quad && \cdots\quad && \bvec{x}_N
\end{array}\right)^\T
\end{eqnarray}
is
\begin{eqnarray}\eqlabel{gp-likelihood}
\ln{p(\bvec{y}\,|\,{X,\,\bvec{\theta}},\,\bvec{\alpha})} =
    -\frac{1}{2} {\bvec{r}_\bvec{\theta}}^\T\,{K_\bvec{\alpha}}^{-1}\,
        \bvec{r}_\bvec{\theta}
    -\frac{1}{2}\ln\det K_\bvec{\alpha}
    - \frac{N}{2} \ln{(2\pi)}
\end{eqnarray}
where
\begin{eqnarray}
    \bvec{r}_\bvec{\theta} = \left(\begin{array}{ccccc}
    y_1 - \mu_\bvec{\theta}(\bvec{x}_1)\quad && \cdots\quad &&
    y_N - \mu_\bvec{\theta}(\bvec{x}_N)
\end{array}\right)^\T
\end{eqnarray}
is the vector of residuals and the elements of the covariance matrix $K$ are
given by $[K_\bvec{\alpha}]_{nm} = k_\bvec{\alpha}(\bvec{x}_n,\,\bvec{x}_m)$.
The maximum likelihood values for the parameters $\bvec{\theta}$ and
$\bvec{\alpha}$ for a given dataset $(\bvec{y},\,X)$ can be found by
maximizing \eq{gp-likelihood} with respect to $\bvec{\theta}$ and
$\bvec{\alpha}$ using a non-linear optimization routine \citep{Nocedal:2006}.
Similarly, probabilistic constraints on $\bvec{\theta}$ and $\bvec{\alpha}$
can be obtained by multiplying the likelihood by a prior
$p(\bvec{\theta},\,\bvec{\alpha})$ and using a Markov Chain Monte Carlo (MCMC)
algorithm to sample the joint posterior probability density.

@@

By rewriting the exponentials in \eq{celerite-kernel-complex} as sums of sine
and cosine functions, we can see the autocorrelation structure is defined by a
mixture of quasiperiodic oscillators
\begin{eqnarray}\eqlabel{celerite-kernel}
k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} +
    \sum_{j=1}^J &&\left[
    a_j\,\exp\left(-c_j\,\tau_{nm}\right)\,\cos\left(d_j\,\tau_{nm}\right)
        \right.\nonumber\\
    &&+ \left.
    b_j\,\exp\left(-c_j\,\tau_{nm}\right)\,\sin\left(d_j\,\tau_{nm}\right)
\right] \quad.
\end{eqnarray}
For clarity, we refer to the argument within the sum as a ``\celeriteterm\
term'' for the remainder of this paper.
The kernel function in \eq{celerite-kernel} is qualitatively similar to the
The Fourier transform\footnote{Here and throughout we have defined the Fourier
transform of the function $f(t)$ as $F(\omega)={(2\,\pi)}^{-1/2}\,
\int_{-\infty}^\infty f(t)\,e^{i\,\omega\,t}\dd t$.} of this covariance
function is the power spectral density (PSD) of the process and it is given by
\begin{eqnarray}\eqlabel{celerite-psd}
S(\omega) = \sum_{j=1}^J \sqrt{\frac{2}{\pi}}
\frac{(a_j\,c_j+b_j\,d_j)\,({c_j}^2+{d_j}^2)+(a_j\,c_j-b_j\,d_j)\,\omega^2}
{\omega^4+2\,({c_j}^2-{d_j}^2)\,\omega^2+{({c_j}^2+{d_j}^2)}^2}\quad.
\end{eqnarray}
The physical interpretation of this model isn't immediately obvious and we
return to a more general discussion shortly but we start by considering some
special cases.

@@

Another source of variability that can be measured from time series
measurements of stars is rotation.
The inhomogeneous surface of the star (spots, plage, \etc) imprints itself as
quasiperiodic variations in photometric or spectroscopic observations
\citep{Dumusque:2014}.
It has been demonstrated that for light curves with nearly uniform sampling,
the empirical autocorrelation function provides a reliable estimate of the
rotation period of a star \citep{Mcquillan:2013, Mcquillan:2014, Aigrain:2015}
and that a GP model with a
quasiperiodic covariance function can be used to make probabilistic
measurements even with sparsely sampled data (R.~Angus, \etal\ submitted).
The covariance function used for this type of analysis has the form
\begin{eqnarray}\eqlabel{sine2}
k(\tau) = A\,\exp\left(-\frac{\tau^2}{2\,\ell^2} -
    \Gamma\,\sin^2\left(\frac{\pi\,\tau}{P} \right) \right)
\end{eqnarray}
where $P$ is the rotation period of the star.
GP modeling with the same kernel function has been proposed as a method of
measuring the mean periodicity in quasiperiodic photometric time series in
general \citep{Wang:2012}.
The key difference between \eq{sine2} and other quasiperiodic kernels is that
it is positive everywhere.
We can construct a simple \celeriteterm\ covariance function with similar properties
as follows
\begin{eqnarray}\eqlabel{rot-kernel}
k(\tau) = \frac{a}{2+b}\,e^{-c\,\tau}\,\left[
    \cos\left(\frac{2\,\pi\,\tau}{P}\right) + (1 + b)
\right]
\end{eqnarray}
for $a>0$, $b>0$, and $c>0$.
The covariance function in \eq{rot-kernel} cannot exactly reproduce \eq{sine2}
but, since \eq{sine2} is only an effective model, \eq{rot-kernel} can be used
as a drop-in replacement for a substantial gain in computational efficiency.

To demonstrate this method, we fit a \celeriteterm\ model with a kernel given by
\eq{rot-kernel} to a \kepler\ light curve for the star KIC~1430163.
This star has a published rotation period of $3.88 \pm 0.58$, measured using
traditional periodogram and autocorrelation function approaches applied to
\kepler\ data from Quarters 0--16 \citep{Mathur:2014}.
We used \emcee\ to sample the joint posterior probability density for the four
parameters in \eq{rot-kernel}, conditioned on two quarters of \kepler\ data,
and find a constraint on the rotation period of $P = \rotationperiod$.
This is in good agreement with the literature value with somewhat smaller
uncertainties.
A detailed comparison of GP rotation period measurements and the traditional
methods is beyond the scope of this paper but forthcoming work (R.~Angus,
\etal, submitted) demonstrates that GP inferences are, at a population level,
more reliable than other methods.
\Figure{rotation} shows a subset of the data used in this example and the
posterior inferences of the PSD and autocorrelation function of the process.
\Figure{rotation-period} shows the marginalized posterior distribution for the
rotation period.

@@

\section{Comparisons to other methods}

@@

For a GP kernel to be valid, it must produce a positive definite covariance
matrix for all input coordinates.
For stationary kernels, this is equivalent~--~by Bochner's theorem \citep[see
Section 4.2.1 in][]{Rasmussen:2006}~--~to requiring that the kernel be the
Fourier transform of a positive finite measure.
This means that the power spectrum of a positive definite kernel must be
positive for all frequencies.
This result is intuitive because, since the power spectrum of a process is
defined as the expected squared amplitude of the Fourier transform of the time
series, it must be non-negative.
